Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-08-19 docs/Manuscript/Manuscript.Rmd
draft rationales section

Clinical decision support (CDS) systems have been a fixture of artificial intelligence (AI) research since inception. As machine learning (ML) emerged, clinical modeling tasks became some of the most common and celebrated applications. A technique core to clinical applications of both AI and ML is _similarity-matching_, which relies on a _patient similarity measure_[^psm] to extract a cohort of past or training cases used to inform the diagnosis, prognosis, or care of a new or test case. Similarity-matching powers the retrieval step of most case-based reasoning (CBR) systems and provides the neighborhoods from which nearest neighbor (NN) models derive predictions. Whereas NN predictions are generated automatically via averaging (regression) or voting (classification), most CBR tools return the retrieved cases to the user to inform human decision-making.

@@

We discuss in this review a hybrid computational approach we call _localized modeling_ that borrows elements from both clinical application areas and appears to have been introduced independently several times. The approach fits ML models, including regularized families of generalized regression models (GLMs), to cohorts retrieved via similarity matching. Whereas CBR and NN prediction use patient similarity to obtain an ideally homogeneous subpopulation matched to the index case, in localized modeling one seeks a relevant yet heterogeneous subpopulation, each member of which deviates slightly from the index case. In principle, this allows interpretable models like GLMs to identify and measure the effects of determining factors that are specific to the index case. In the reviewed studies, localized modeling shows some promise toward simultaneously achieving the performance improvements of ML and the interpretability of classical models.

@@

We refer the interested reader to several previous reviews of CBR in medicine [@Gierl1998; @Begum2011; @Choudhury2016], of measures of patient similarity [@Dai2020], and of uses of patient similarity in predictive models [@Welch2013; @Sharafoddini2017; @Parimbelli2018]. While the reviews of CBR focus on applications using health data, the patient similarity reviews encompass many additional types of data (various molecular -omics, genetic tests, medical images, laboratory tests, patient preferences, patient-reported outcomes, tracking devices, social media) and survey a much broader scope of models (exploratory analysis / dimension reduction / cluster analysis, risk evaluation / outcome prediction, clinical decision support / software tools). For the present review, we are interested in how similarity matching on patient-level health data is used to construct local cohorts for predictive modeling.

@@

In Covidence, two authors (AC and JCB) independently evaluated each manuscript for the four eligibility criteria. The first criterion that a manuscript failed was designated the reason for exclusion. An antecedent criterion was used to exclude manuscripts that did not report the results of original studies involving real-world experiments or empirical data, for example surveys of prior work and proposals of frameworks. In cases of disagreement between the authors on the reason for exclusion, the first criterion was adopted. There were 60 disagreements over inclusion versus exclusion. The authors arrived at agreement on each manuscript through discussion. We calculated inter-rater reliability to evaluate our screening and review process.

@@

Finally, one author (JCB) applied the same review process to the seed set of 43 studies that inspired the review. This obtained 6 additional studies that had not been retrieved by any of our database searches [@Park2006; @Lowsky2013; @Lee2015; @Ng2015; @Lee2017; @Wang2019].

@@

We then reviewed 43 studies comprising a seed set that inspired this review. After removing duplicates and screening for eligibility, we were left with 6 additional studies, 1 of which was excluded from the synthesis for using nearest neighbors prediction. Finally, citations in reference lists from the 43 + 328 = 371 studies assessed for eligibility led us to identify 1 additional study that met our criteria [@Kasabov2010]. This left us with 19 + 1 + 5 = 25 studies included in the review and synthesis.
Inter-rater reliability was low, at 82% relative to a 72% probability of random agreement.

@@

Another dominant characteristic of included studies is their remit. We categorized studies as "knowledge" or "practice" based on whether their aim was to produce generalizable knowledge or to improve practice. We classified 7 of the 25 studies as "knowledge", making the majority of studies from a "practice" standpoint. These studies aimed primarily to provide providers with tools to use in clinic or to improve outcomes for patients currently receiving care.

@@

In addition to their remit, we coded several aspects of the design of each study, including the source and type of data and the clinical task the model performed (Table \ref{tab:synthesis}), as well as the specific methodology and terminology adopted (Table \ref{tab:composite}).
In the following subsections, we take a closer look at these design elements.

@@

## Application domain

While all included studies were reported in scientific and medical journals, the vast majority were oriented toward clinical practice rather than medical research. For example, a 2006 study specifically evaluated the usefulness of CBR-based explanations for the purpose of decision support [@Doyle2006a]. This study was part of a much larger literature on CBR systems and was included here despite relying on NN prediction because it used a non-traditional voting scheme to generate recommendations. A more recent study took essentially the same focus with respect to a proposed clinical risk prediction model, which amounted to CBR with a novel weighting scheme on predictors informed by expert consensus [@Fang2021a]. In both cases a prototype implementation was deployed in an experimental setting for evaluation.

@@

The mosaic plot in Figure \ref{fig:properties} summarizes the relation between data source, clinical task, and remit.

@@

@@

@@

### Coherence

@@

Distinct from but related to similarity learning was supervised cohort construction. This was an adaptive step taken by @Park2006 to optimize the predictive performance of models fitted to cohorts retrieved using a fixed similarity rather than count threshold, a counterpart to conventional CBR they called "statistical CBR". The step used a heuristic procedure to locate a similarity threshold that (locally) maximizes predictive accuracy on the training set. This is analogous to optimizing the neighborhood size parameter in NN predictive modeling, so we will not discuss it further except to mention that the authors found statistical CBR to outperform conventional CBR on several data sets. @CampilloGimenez2012a combined these learning techniques: They used logistic regression on the training set to obtain weights for the predictors (by predicting the binary outcome of kidney transplant waitlist registration) and for the cases (by predicting agreement of outcome between an index case and other training cases), and they used exhaustion to optimize the size of the retrieved cohort for the accuracy of the NN model using the previously optimized weights.

@@

Each step can be unsupervised supervised: Unsupervised similarity measures include discrete measures like the Levenshtein distance and continuous measures like cosine similarity, while supervised measures include the Mahalanobis distance and random forest proximity as well as several composite measures with weights calculated from the data. Most retrieval steps obtained cohorts with a uniform size (cardinality) or similarity threshold, and the remaining were likewise only informed by predictors; thus, while in principle a retrieval step could be supervised, in this sample none were. Almost all models were supervised, being that most performed predictive tasks, though @Mariuzzi1997 modeled only survival rates within localized cohorts. Meanwhile, each step can be optimized in a ML fashion by having its parameters tuned to improve performance. We found no studies that tuned the calculation of similarity, though most uniform cohort sizes were tuned. Some earlier studies tuned generalized regression models, but more recent studies have not.

@@

@@

* Search results, included studies, and other bibliography: <https://www.zotero.org/groups/5017571/imsr/>
* Bibliographic and methodological properties of included studies: <https://docs.google.com/spreadsheets/d/1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo/>
* Terminology and composite techniques of included studies: <https://docs.google.com/spreadsheets/d/1xvDJwiLBoI2oz8fxHJ5MjNmiju_RAlK7RJv-wXe1DAs/>
* Code used to perform analyses and prepare the manuscript: <https://github.com/corybrunson/imsr>

@@

Following discussion among AC, PMJ, and JCB, we omitted results from Google Scholar due to missing abstracts, missing URLs, high overlap with other search results, and irreproducibility of the search process.

@@

>>>>

Clinical decision support (CDS) systems have been a fixture of artificial intelligence (AI) research since inception. As machine learning (ML) emerged, clinical modeling tasks became some of the most common and celebrated applications. A technique core to clinical applications of both AI and ML is _similarity-matching_, which relies on a _patient similarity measure_[^psm] to extract a cohort of past or training cases used to inform the diagnosis, prognosis, or care of a new or test case. Similarity-matching powers the retrieval step of most case-based reasoning (CBR) systems and provides the neighborhoods from which nearest neighbor (NN) models derive predictions. Whereas NN predictions are generated automatically via averaging (regression) or voting (classification), traditional CBR tools return the retrieved cases to the user to inform human decision-making.

@@

We discuss in this review a hybrid computational approach we call _localized modeling_ that borrows elements from both clinical application areas and appears to have been introduced independently several times. The approach fits ML models, including regularized families of generalized regression models (GLMs), to cohorts retrieved via similarity matching. Whereas CBR and NN prediction use patient similarity to obtain an ideally homogeneous subpopulation matched to the index case, in localized modeling one seeks a relevant yet heterogeneous subpopulation, each member of which deviates slightly from the index case. In principle, this allows interpretable models like GLMs to identify and measure the effects of determining factors that are specific to the index case.
The use of similarity cohorts distinguishes localized modeling from classical GLMs that still dominate analysis in clinical trials and comparative effectiveness research, as well as from many cutting-edge ML models that are fitted to whole populations, while the cohorts' heterogeneity and its utility to the model distinguish the approach from CBR.
In the reviewed studies, localized modeling shows some promise toward simultaneously achieving the performance improvements of ML and the interpretability of classical models.

<!--
In multivariate settings, because similarity decreases with difference along each factor, included cases that differ more with respect to some factors will tend to differ less with respect to others. Consequently, (1) the distribution of a factor in the subpopulation will tend to be centered near its value at the index case, and (2) the joint distribution of a pair of factors will tend to be sparse away from both center axes, i.e. when both factors differ from their values at the index case. This should produce localized models that are less responsive to deviation along multiple uncorrelated factors, so that the influence of each factor in the model reflects more its independent effect than its interactions with other factors.
-->

@@

We refer the interested reader to several previous reviews of CBR in medicine [@Gierl1998; @Begum2011; @Choudhury2016], of measures of patient similarity [@Dai2020], and of uses of patient similarity in predictive models [@Welch2013; @Sharafoddini2017; @Parimbelli2018]. While the reviews of CBR focus on applications using health data, the patient similarity reviews encompass many additional types of data (various molecular -omics, genetic tests, medical images, laboratory tests, patient preferences, patient-reported outcomes, tracking devices, social media) and survey a much broader scope of models (exploratory analysis via dimension reduction and cluster analysis; risk evaluation and outcome prediction; clinical decision support and software tools). For the present review, we are interested in how similarity matching on patient-level health data is used to construct local cohorts for predictive modeling.

@@

In Covidence, two authors (AC and JCB) independently evaluated each manuscript for the four eligibility criteria. The first criterion that a manuscript failed was designated the reason for exclusion. An antecedent criterion was used to exclude manuscripts that did not report the results of original studies involving real-world experiments or empirical data, for example surveys of prior work and proposals of frameworks. In cases of disagreement between the authors on the reason for exclusion, the first criterion was adopted. The authors arrived at agreement on inclusion or exclusion through discussion. We calculated inter-rater reliability to evaluate our screening and review process.

@@

Finally, one author (JCB) applied the same review process to the seed set of 43 studies that inspired the review.

@@

There were 60 disagreements over inclusion versus exclusion.
Inter-rater reliability was low, at 82% relative to a 72% probability of random agreement.

We then reviewed 43 studies comprising a seed set that inspired this review. After removing duplicates and screening for eligibility, we were left with 6 additional studies [@Park2006; @Lowsky2013; @Lee2015; @Ng2015; @Lee2017; @Wang2019], 1 of which was excluded from the synthesis for using nearest neighbors prediction. Finally, citations in reference lists from the 43 + 328 = 371 studies assessed for eligibility led us to identify 1 additional study that met our criteria [@Kasabov2010]. This left us with 19 + 1 + 5 = 25 studies included in the review and synthesis.

@@

Another dominant characteristic of included studies is their broader remit. We categorized studies as "knowledge" or "practice" based on whether their aim was to produce generalizable knowledge or to improve practice. We classified 7 of the 25 studies as "knowledge", making the majority of studies from a "practice" standpoint. These studies aimed primarily to provide providers with tools to use in clinic or to improve outcomes for patients currently receiving care.

@@

In addition to the aim of its analysis, we coded several aspects of the design of each study, including the source and type of data and the clinical task the model performed (Table \ref{tab:synthesis}), as well as the specific methodology and terminology adopted (Table \ref{tab:composite}).
In the following subsections, we take a closer look at these design elements.

@@

## Application domains

While all included studies were reported in scientific and medical journals, the vast majority were oriented toward clinical practice rather than medical research. For example, a 2006 study specifically evaluated the usefulness of CBR-based explanations for the purpose of decision support [@Doyle2006a]. This study was part of a much larger literature on CBR systems and was included here despite relying on NN prediction because it used an unconventional voting scheme to generate recommendations. A more recent study took essentially the same focus with respect to a proposed clinical risk prediction model, which amounted to CBR with a novel weighting scheme on predictors informed by expert consensus [@Fang2021a]. In both cases a prototype implementation was deployed in an experimental setting for evaluation.

@@

The mosaic plot in Figure \ref{fig:properties} summarizes the relation between data source, clinical task, and aim.

@@

## Rationales

The included studies hypothesized, asserted, or assumed several benefits of localized modeling specific to clinical and medical settings and advantages over other modeling approaches (see Section \ref{sec:availability}).
The most common was that the restriction to similar or relevant past cases would improve predictive performance for the index case [@Mariuzzi1997; @Liang2015a; @Ng2015; @Lee2017]. In particular, @Lee2015 hypothesized and confirmed that the value of each past case would be positively related to its similarity to the index case, an assumption built in to the weighting schemes of other approaches. @Lowsky2013 made a different case, that the fewer parametric assumptions and complications of a CBR-style model would allow for greater accuracy.
Additionally, several investigators asserted that the use of localized cohorts befit the clinical focus on the individual patient rather than the population, without reference to model performance [@Song2006c; @Xu2008a].

In an interesting contrast, @Tang2021c and @Ng2021b argued that their localized approach using the larger and more heterogeneous populations covered by EHR-derived data could better capture the messy and diverse lessons of everyday practice, as a counterpart to guidelines based on randomized controlled trials.
Their emphasis on the enabling role of EHRs to power methods well-suited to large, structured data repositories was shared by several others [@CampilloGimenez2012a; @Nicolas2014a; @Verma2015a; @Lee2017].
@Verma2015a and @Zhang2018a additionally pointed out that localized models are adaptable to noise in the data, variation in patterns of missingness, and (harkening to the last step of the R4 cycle) addition of new cases with known outcomes to the corpus. These properties, they said, tend to be more difficult for whole-population models to handle.

The other frequent advantage attributed to localized modeling was interpretability.
@Elter2007a and @Nicolas2014a emphasized the value of the detailed past cases, available to the user, on which predictions are based. This "self-explanation capability" made outputs more intelligible to physicians in the CDS setting.
@Ng2015 additional pointed out that localized GLMs yield localized effect estimates, which may help investigators identify individually relevant risk factors.

The remaining coded rationales were for augmentations or hybridizations of then-conventional CBR, most of which defended the use of other tools to improve performance via cohort selection [@CampilloGimenez2012a; @Nicolas2014a; @Vilhena2016a] or to make models and outputs more interpretable [@Lopez2011a; @Wang2019]. @Liang2015a argued for simultaneous optimization of cohort and feature selection with model parameterization; their TWNFI approach combines localization with regularization.

@@

**Move part of this paragraph to Discussion.**

@@

## Coherence

@@

Distinct from but related to similarity learning was supervised cohort construction. This was an adaptive step taken by @Park2006 to optimize the predictive performance of models fitted to cohorts retrieved using a fixed similarity rather than count threshold, a counterpart to conventional CBR they called "statistical CBR".[^combinatorial] The step used a heuristic procedure to locate a similarity threshold that (locally) maximizes predictive accuracy on the training set. This is analogous to optimizing the neighborhood size parameter in NN predictive modeling, so we will not discuss it further except to mention that the authors found statistical CBR to outperform conventional CBR on several data sets. @CampilloGimenez2012a combined these learning techniques: They used logistic regression on the training set to obtain weights for the predictors (by predicting the binary outcome of kidney transplant waitlist registration) and for the cases (by predicting agreement of outcome between an index case and other training cases), and they used exhaustion to optimize the size of the retrieved cohort for the accuracy of the NN model using the previously optimized weights.

[^combinatorial]: Drawing from @Goyal2008, we suggest the term "combinatorial CBR" for the then-conventional approach using similarity cohorts of fixed cardinality.

@@

Each step can be unsupervised supervised: Unsupervised similarity measures include discrete measures like the Levenshtein distance and continuous measures like cosine similarity, while supervised measures include the Mahalanobis distance and random forest proximity as well as several composite measures with weights calculated from the data. Most retrieval steps obtained cohorts with a uniform size (cardinality) or similarity threshold, and the remaining were likewise only informed by predictors; thus, while a retrieval step could in principle be supervised, in this sample none were. Almost all models were supervised, being that most performed predictive tasks, though @Mariuzzi1997 modeled only survival rates within localized cohorts. Meanwhile, each step can be optimized in a ML fashion by having its parameters tuned to improve performance. We found no studies that tuned the calculation of similarity, though most uniform cohort sizes were tuned. Some earlier studies tuned generalized regression models, but more recent studies have not.

@@

We are grateful for the astute and supportive comments of three anonymous reviewers.

@@

* Search results, included studies, and other bibliography: <https://www.zotero.org/groups/5017571/imsr/>
* Bibliographic and methodological properties of included studies: <https://docs.google.com/spreadsheets/d/1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo/>
* Terminology and composite techniques of included studies: <https://docs.google.com/spreadsheets/d/1xvDJwiLBoI2oz8fxHJ5MjNmiju_RAlK7RJv-wXe1DAs/>
* Code used to conduct analyses and prepare the manuscript: <https://github.com/corybrunson/imsr>

@@

Following discussion among AC, PMJ, and JCB, we discarded results from Google Scholar due to missing abstracts, missing URLs, high overlap with other search results, and irreproducibility of the search process.

@@

**Write this section!**

### 2023-08-19 docs/Manuscript/Manuscript.tex
draft rationales section

Clinical decision support (CDS) systems have been a fixture of
artificial intelligence (AI) research since inception. As machine
learning (ML) emerged, clinical modeling tasks became some of the most
common and celebrated applications. A technique core to clinical
applications of both AI and ML is \emph{similarity-matching}, which
relies on a \emph{patient similarity measure}\footnote{Similarity
  measures between more granular units of analysis, such as encounters
  and decision points, are often still referred to as patient similarity
  measures.} to extract a cohort of past or training cases used to
inform the diagnosis, prognosis, or care of a new or test case.
Similarity-matching powers the retrieval step of most case-based
reasoning (CBR) systems and provides the neighborhoods from which
nearest neighbor (NN) models derive predictions. Whereas NN predictions
are generated automatically via averaging (regression) or voting
(classification), most CBR tools return the retrieved cases to the user
to inform human decision-making.

We discuss in this review a hybrid computational approach we call
\emph{localized modeling} that borrows elements from both clinical
application areas and appears to have been introduced independently
several times. The approach fits ML models, including regularized
families of generalized regression models (GLMs), to cohorts retrieved
via similarity matching. Whereas CBR and NN prediction use patient
similarity to obtain an ideally homogeneous subpopulation matched to the
index case, in localized modeling one seeks a relevant yet heterogeneous
subpopulation, each member of which deviates slightly from the index
case. In principle, this allows interpretable models like GLMs to
identify and measure the effects of determining factors that are
specific to the index case. In the reviewed studies, localized modeling
shows some promise toward simultaneously achieving the performance
improvements of ML and the interpretability of classical models.

@@

We refer the interested reader to several previous reviews of CBR in
medicine {[}3,5,6{]}, of measures of patient similarity {[}7{]}, and of
uses of patient similarity in predictive models {[}8--10{]}. While the
reviews of CBR focus on applications using health data, the patient
similarity reviews encompass many additional types of data (various
molecular -omics, genetic tests, medical images, laboratory tests,
patient preferences, patient-reported outcomes, tracking devices, social
media) and survey a much broader scope of models (exploratory analysis /
dimension reduction / cluster analysis, risk evaluation / outcome
prediction, clinical decision support / software tools). For the present
review, we are interested in how similarity matching on patient-level
health data is used to construct local cohorts for predictive modeling.

@@

Finally, one author (JCB) applied the same review process to the seed
set of 43 studies that inspired the review. This obtained 6 additional
studies that had not been retrieved by any of our database searches
{[}12--17{]}.

@@

We then reviewed 43 studies comprising a seed set that inspired this
review. After removing duplicates and screening for eligibility, we were
left with 6 additional studies, 1 of which was excluded from the
synthesis for using nearest neighbors prediction. Finally, citations in
reference lists from the 43 + 328 = 371 studies assessed for eligibility
led us to identify 1 additional study that met our criteria {[}18{]}.
This left us with 19 + 1 + 5 = 25 studies included in the review and
synthesis. Inter-rater reliability was low, at 82\% relative to a 72\%
probability of random agreement.

@@

Another dominant characteristic of included studies is their remit. We
categorized studies as ``knowledge'' or ``practice'' based on whether
their aim was to produce generalizable knowledge or to improve practice.
We classified 7 of the 25 studies as ``knowledge'', making the majority
of studies from a ``practice'' standpoint. These studies aimed primarily
to provide providers with tools to use in clinic or to improve outcomes
for patients currently receiving care.

@@

\hypertarget{application-domain}{%
\subsection{Application domain}\label{application-domain}}

While all included studies were reported in scientific and medical
journals, the vast majority were oriented toward clinical practice
rather than medical research. For example, a 2006 study specifically
evaluated the usefulness of CBR-based explanations for the purpose of
decision support {[}36{]}. This study was part of a much larger
literature on CBR systems and was included here despite relying on NN
prediction because it used a non-traditional voting scheme to generate
recommendations. A more recent study took essentially the same focus
with respect to a proposed clinical risk prediction model, which
amounted to CBR with a novel weighting scheme on predictors informed by
expert consensus {[}37{]}. In both cases a prototype implementation was
deployed in an experimental setting for evaluation.

@@

@@

@@

\hypertarget{coherence}{%
\subsubsection{Coherence}\label{coherence}}

@@

Distinct from but related to similarity learning was supervised cohort
construction. This was an adaptive step taken by {[}13{]} to optimize
the predictive performance of models fitted to cohorts retrieved using a
fixed similarity rather than count threshold, a counterpart to
conventional CBR they called ``statistical CBR''. The step used a
heuristic procedure to locate a similarity threshold that (locally)
maximizes predictive accuracy on the training set. This is analogous to
optimizing the neighborhood size parameter in NN predictive modeling, so
we will not discuss it further except to mention that the authors found
statistical CBR to outperform conventional CBR on several data sets.
{[}27{]} combined these learning techniques: They used logistic
regression on the training set to obtain weights for the predictors (by
predicting the binary outcome of kidney transplant waitlist
registration) and for the cases (by predicting agreement of outcome
between an index case and other training cases), and they used
exhaustion to optimize the size of the retrieved cohort for the accuracy
of the NN model using the previously optimized weights.

@@

Each step can be unsupervised supervised: Unsupervised similarity
measures include discrete measures like the Levenshtein distance and
continuous measures like cosine similarity, while supervised measures
include the Mahalanobis distance and random forest proximity as well as
several composite measures with weights calculated from the data. Most
retrieval steps obtained cohorts with a uniform size (cardinality) or
similarity threshold, and the remaining were likewise only informed by
predictors; thus, while in principle a retrieval step could be
supervised, in this sample none were. Almost all models were supervised,
being that most performed predictive tasks, though {[}11{]} modeled only
survival rates within localized cohorts. Meanwhile, each step can be
optimized in a ML fashion by having its parameters tuned to improve
performance. We found no studies that tuned the calculation of
similarity, though most uniform cohort sizes were tuned. Some earlier
studies tuned generalized regression models, but more recent studies
have not.

@@

\begin{itemize}
\tightlist
\item
  Search results, included studies, and other bibliography:
  \url{https://www.zotero.org/groups/5017571/imsr/}
\item
  Bibliographic and methodological properties of included studies:
  \url{https://docs.google.com/spreadsheets/d/1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo/}
\item
  Terminology and composite techniques of included studies:
  \url{https://docs.google.com/spreadsheets/d/1xvDJwiLBoI2oz8fxHJ5MjNmiju_RAlK7RJv-wXe1DAs/}
\item
  Code used to perform analyses and prepare the manuscript:
  \url{https://github.com/corybrunson/imsr}
\end{itemize}

@@

@@

>>>>

Clinical decision support (CDS) systems have been a fixture of
artificial intelligence (AI) research since inception. As machine
learning (ML) emerged, clinical modeling tasks became some of the most
common and celebrated applications. A technique core to clinical
applications of both AI and ML is \emph{similarity-matching}, which
relies on a \emph{patient similarity measure}\footnote{Similarity
  measures between more granular units of analysis, such as encounters
  and decision points, are often still referred to as patient similarity
  measures.} to extract a cohort of past or training cases used to
inform the diagnosis, prognosis, or care of a new or test case.
Similarity-matching powers the retrieval step of most case-based
reasoning (CBR) systems and provides the neighborhoods from which
nearest neighbor (NN) models derive predictions. Whereas NN predictions
are generated automatically via averaging (regression) or voting
(classification), traditional CBR tools return the retrieved cases to
the user to inform human decision-making.

We discuss in this review a hybrid computational approach we call
\emph{localized modeling} that borrows elements from both clinical
application areas and appears to have been introduced independently
several times. The approach fits ML models, including regularized
families of generalized regression models (GLMs), to cohorts retrieved
via similarity matching. Whereas CBR and NN prediction use patient
similarity to obtain an ideally homogeneous subpopulation matched to the
index case, in localized modeling one seeks a relevant yet heterogeneous
subpopulation, each member of which deviates slightly from the index
case. In principle, this allows interpretable models like GLMs to
identify and measure the effects of determining factors that are
specific to the index case. The use of similarity cohorts distinguishes
localized modeling from classical GLMs that still dominate analysis in
clinical trials and comparative effectiveness research, as well as from
many cutting-edge ML models that are fitted to whole populations, while
the cohorts' heterogeneity and its utility to the model distinguish the
approach from CBR. In the reviewed studies, localized modeling shows
some promise toward simultaneously achieving the performance
improvements of ML and the interpretability of classical models.

@@

We refer the interested reader to several previous reviews of CBR in
medicine {[}3,5,6{]}, of measures of patient similarity {[}7{]}, and of
uses of patient similarity in predictive models {[}8--10{]}. While the
reviews of CBR focus on applications using health data, the patient
similarity reviews encompass many additional types of data (various
molecular -omics, genetic tests, medical images, laboratory tests,
patient preferences, patient-reported outcomes, tracking devices, social
media) and survey a much broader scope of models (exploratory analysis
via dimension reduction and cluster analysis; risk evaluation and
outcome prediction; clinical decision support and software tools). For
the present review, we are interested in how similarity matching on
patient-level health data is used to construct local cohorts for
predictive modeling.

@@

Finally, one author (JCB) applied the same review process to the seed
set of 43 studies that inspired the review.

@@

There were 60 disagreements over inclusion versus exclusion. Inter-rater
reliability was low, at 82\% relative to a 72\% probability of random
agreement.

We then reviewed 43 studies comprising a seed set that inspired this
review. After removing duplicates and screening for eligibility, we were
left with 6 additional studies {[}12--17{]}, 1 of which was excluded
from the synthesis for using nearest neighbors prediction. Finally,
citations in reference lists from the 43 + 328 = 371 studies assessed
for eligibility led us to identify 1 additional study that met our
criteria {[}18{]}. This left us with 19 + 1 + 5 = 25 studies included in
the review and synthesis.

@@

Another dominant characteristic of included studies is their broader
remit. We categorized studies as ``knowledge'' or ``practice'' based on
whether their aim was to produce generalizable knowledge or to improve
practice. We classified 7 of the 25 studies as ``knowledge'', making the
majority of studies from a ``practice'' standpoint. These studies aimed
primarily to provide providers with tools to use in clinic or to improve
outcomes for patients currently receiving care.

@@

\hypertarget{application-domains}{%
\subsection{Application domains}\label{application-domains}}

While all included studies were reported in scientific and medical
journals, the vast majority were oriented toward clinical practice
rather than medical research. For example, a 2006 study specifically
evaluated the usefulness of CBR-based explanations for the purpose of
decision support {[}36{]}. This study was part of a much larger
literature on CBR systems and was included here despite relying on NN
prediction because it used an unconventional voting scheme to generate
recommendations. A more recent study took essentially the same focus
with respect to a proposed clinical risk prediction model, which
amounted to CBR with a novel weighting scheme on predictors informed by
expert consensus {[}37{]}. In both cases a prototype implementation was
deployed in an experimental setting for evaluation.

@@

\hypertarget{rationales}{%
\subsection{Rationales}\label{rationales}}

The included studies hypothesized, asserted, or assumed several benefits
of localized modeling specific to clinical and medical settings and
advantages over other modeling approaches (see Section
\ref{sec:availability}). The most common was that the restriction to
similar or relevant past cases would improve predictive performance for
the index case {[}11,15,16,26{]}. In particular, {[}14{]} hypothesized
and confirmed that the value of each past case would be positively
related to its similarity to the index case, an assumption built in to
the weighting schemes of other approaches. {[}12{]} made a different
case, that the fewer parametric assumptions and complications of a
CBR-style model would allow for greater accuracy. Additionally, several
investigators asserted that the use of localized cohorts befit the
clinical focus on the individual patient rather than the population,
without reference to model performance {[}21,23{]}.

In an interesting contrast, {[}34{]} and {[}35{]} argued that their
localized approach using the larger and more heterogeneous populations
covered by EHR-derived data could better capture the messy and diverse
lessons of everyday practice, as a counterpart to guidelines based on
randomized controlled trials. Their emphasis on the enabling role of
EHRs to power methods well-suited to large, structured data repositories
was shared by several others {[}16,25,27,28{]}. {[}25{]} and {[}30{]}
additionally pointed out that localized models are adaptable to noise in
the data, variation in patterns of missingness, and (harkening to the
last step of the R4 cycle) addition of new cases with known outcomes to
the corpus. These properties, they said, tend to be more difficult for
whole-population models to handle.

The other frequent advantage attributed to localized modeling was
interpretability. {[}22{]} and {[}28{]} emphasized the value of the
detailed past cases, available to the user, on which predictions are
based. This ``self-explanation capability'' made outputs more
intelligible to physicians in the CDS setting. {[}15{]} additional
pointed out that localized GLMs yield localized effect estimates, which
may help investigators identify individually relevant risk factors.

The remaining coded rationales were for augmentations or hybridizations
of then-conventional CBR, most of which defended the use of other tools
to improve performance via cohort selection {[}27--29{]} or to make
models and outputs more interpretable {[}17,24{]}. {[}26{]} argued for
simultaneous optimization of cohort and feature selection with model
parameterization; their TWNFI approach combines localization with
regularization.

@@

\textbf{Move part of this paragraph to Discussion.}

@@

\hypertarget{coherence}{%
\subsection{Coherence}\label{coherence}}

@@

Distinct from but related to similarity learning was supervised cohort
construction. This was an adaptive step taken by {[}13{]} to optimize
the predictive performance of models fitted to cohorts retrieved using a
fixed similarity rather than count threshold, a counterpart to
conventional CBR they called ``statistical CBR''.\footnote{Drawing from
  {[}42{]}, we suggest the term ``combinatorial CBR'' for the
  then-conventional approach using similarity cohorts of fixed
  cardinality.} The step used a heuristic procedure to locate a
similarity threshold that (locally) maximizes predictive accuracy on the
training set. This is analogous to optimizing the neighborhood size
parameter in NN predictive modeling, so we will not discuss it further
except to mention that the authors found statistical CBR to outperform
conventional CBR on several data sets. {[}27{]} combined these learning
techniques: They used logistic regression on the training set to obtain
weights for the predictors (by predicting the binary outcome of kidney
transplant waitlist registration) and for the cases (by predicting
agreement of outcome between an index case and other training cases),
and they used exhaustion to optimize the size of the retrieved cohort
for the accuracy of the NN model using the previously optimized weights.

@@

Each step can be unsupervised supervised: Unsupervised similarity
measures include discrete measures like the Levenshtein distance and
continuous measures like cosine similarity, while supervised measures
include the Mahalanobis distance and random forest proximity as well as
several composite measures with weights calculated from the data. Most
retrieval steps obtained cohorts with a uniform size (cardinality) or
similarity threshold, and the remaining were likewise only informed by
predictors; thus, while a retrieval step could in principle be
supervised, in this sample none were. Almost all models were supervised,
being that most performed predictive tasks, though {[}11{]} modeled only
survival rates within localized cohorts. Meanwhile, each step can be
optimized in a ML fashion by having its parameters tuned to improve
performance. We found no studies that tuned the calculation of
similarity, though most uniform cohort sizes were tuned. Some earlier
studies tuned generalized regression models, but more recent studies
have not.

@@

\begin{itemize}
\tightlist
\item
  Search results, included studies, and other bibliography:
  \url{https://www.zotero.org/groups/5017571/imsr/}
\item
  Bibliographic and methodological properties of included studies:
  \url{https://docs.google.com/spreadsheets/d/1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo/}
\item
  Terminology and composite techniques of included studies:
  \url{https://docs.google.com/spreadsheets/d/1xvDJwiLBoI2oz8fxHJ5MjNmiju_RAlK7RJv-wXe1DAs/}
\item
  Code used to conduct analyses and prepare the manuscript:
  \url{https://github.com/corybrunson/imsr}
\end{itemize}

@@

\textbf{Write this section!}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Goyal2008}{}}%
\CSLLeftMargin{{[}42{]} }%
\CSLRightInline{Goyal N, Lifshits Y, Schütze H. Disorder inequality: A
combinatorial approach to nearest neighbor search. Proceedings of the
2008 {International Conference} on {Web Search} and {Data Mining}, {New
York, NY, USA}: {Association for Computing Machinery}; 2008, p. 25--32.
\url{https://doi.org/10.1145/1341531.1341538}.}
